Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/intro.tex
complete intro

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
\end{lstlisting}
%
This scenario, in which a multivariate function is applied many times with one
argument fixed, arises frequently. 
\crem{?? Many algorithm design techniques revolve around the idea of finding the
computations in $f$ that depend only on $x$, and
performing them only once.
\\
?? It's advantageous to find the computations in f that depend on x and perform
them only once.}
This is the intent of compiler optimization techniques like loop hoisting and
common subexpression elimination, which identify the computations common to all
invocations of $f$, and perform them once at the beginning of the computation.

Programs structured in this way are particularly important in real-time computer
graphics; in fact, modern graphics architectures \emph{require} that graphics
computations be organized into distinct {\em passes} which perform increasingly
fine-grained computations (for example, per object, per region of the screen,
and per pixel, where later computations use the results from the prior
computations).

While performant, this paradigm of programming with explicit collections of
passes results in complex code where invariants must hold across different
passes, and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

>>>>

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
\end{lstlisting}
%
In this common scenario, a multivariate function is applied many times with one
argument fixed, so it is advantageous to find the computations in $f$ that
depend on $x$ and perform them only once. This is the intent of compiler
optimization techniques like loop hoisting and common subexpression elimination,
which identify the computations common to all invocations of $f$, and perform
them once at the beginning of the computation.

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass).

While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.
